---
title: "birds_dash"
author: "Cody"
date: "2024-05-19"
output: html_document
---

# Import packages 
```{r}
library(arrow)
library(tidyverse)
library(magrittr)
library(ggpubr)
library(stringr)
library(rlang)
library(ggplot2); theme_set(theme_bw())
library(googlesheets4)
library(fdapace)
library(dplyr)
library(plyr)
library(pracma)
library(tidyr)
library(kdensity)
library(circular)
library(fda)
library(gridExtra)
library(patchwork)

```

# Clear Environment? 

```{r}
rm(list = ls())
```


# Import Raw Counts Data

```{r}

#raw counts data
ebird19counts = read_parquet("~/Desktop/repos/birds_dash/data/eb19_raw_counts_all.parquet")
ebird22counts = read_parquet("~/Desktop/repos/birds_dash/data/eb22_raw_counts_all.parquet")
inat19counts = read_parquet("~/Desktop/repos/birds_dash/data/in19_raw_counts_all.parquet")
inat22counts = read_parquet("~/Desktop/repos/birds_dash/data/in22_raw_counts_all.parquet")

birdids_df = read.csv("~/Desktop/repos/birds_dash/data/birdids.csv")

birdids = birdids_df$x

#dictionary & traits
bird_dict = read.csv("~/Desktop/repos/explorations_func_data/name_aviD_dict_22.csv")
trait_data = read.csv("~/Desktop/bird_data/ELEData/TraitData/AVONET1_BirdLife.csv")

#process trait data
#isolate avibaseID
split_list = str_split(trait_data$Avibase.ID1, "-")
avibaseID_list = lapply(split_list, function(x) x[2])
trait_data$AVIBASE = unlist(avibaseID_list)


```


# Filter to 260 Birds

```{r}
#filter to 260 birds:
#(How do we decide these 260 birds? Write up.) 
ebird19counts = ebird19counts %>% filter(AVIBASEID %in% birdids)
inat19counts = inat19counts %>% filter(AVIBASEID %in% birdids)
ebird22counts = ebird22counts %>% filter(AVIBASEID %in% birdids)
inat22counts = inat22counts %>% filter(AVIBASEID %in% birdids)
```


#Combine into 1 df and also update separated versions with db/yr key

```{r}
count_list = list(ebird19 = ebird19counts,
                  inat19 = inat19counts, 
                  ebird22 = ebird22counts, 
                  inat22 = inat22counts)
allcounts = bind_rows(count_list, .id = "yr_db")


ebird19counts = allcounts %>% filter(yr_db == "ebird19")
ebird22counts = allcounts %>% filter(yr_db == "ebird22")
inat19counts = allcounts %>% filter(yr_db == "inat19")
inat22counts = allcounts %>% filter(yr_db == "inat22")

```



```{r}

counts_to_relative = function(df){
  df_relative = df

  for(i in 1:nrow(df)){
    df_relative[i, 2:53] = df[i, 2:53]/sum(df[i, 2:53]) #divide row #s by total to get relative counts
  }

  return(df_relative)
#sanity check that they add to 1
#df_relative[,2:53] %>% rowSums()
}

ebird19rel = counts_to_relative(ebird19counts)
ebird22rel = counts_to_relative(ebird22counts)
inat19rel = counts_to_relative(inat19counts)
inat22rel = counts_to_relative(inat22counts)



```

#save unsmoothed relative data
```{r}
# write.csv(ebird19rel, "~/Desktop/repos/birds_dash/data/ebird19rel_raw.csv")
# write.csv(ebird22rel, "~/Desktop/repos/birds_dash/data/ebird22rel_raw.csv")
# write.csv(inat19rel, "~/Desktop/repos/birds_dash/data/inat19rel_raw.csv")
# write.csv(inat19rel, "~/Desktop/repos/birds_dash/data/inat22rel_raw.csv")
```


#fourier smoothing
```{r}
four_smooth = function(df_relative, k = 3){
  #establish time grid
  t = 1:52
  #curve data
  database = df_relative[,1]
  avibaseid = df_relative[,54]
  y = unlist(t(df_relative[,2:53]))
  wk_names = colnames(df_relative[,2:53])
  #create a basis of dimension k 
  k_basis = create.fourier.basis(rangeval = range(t), 
                                 nbasis = k)
  #smooth the curve data using the dimension of basis k 
  smooth_y_basis = smooth.basis(argvals = t, 
                                y = y, 
                                fdParobj = k_basis)$fd
  eval_y = eval.fd(t, smooth_y_basis)
  
  #prevent negatives
  eval_y = pmax(eval_y, 0.001) 
  
  #renormalize
  dens = eval_y
  for(i in 1:ncol(eval_y)){
    dens[,i] = eval_y[,i] / trapz(t, eval_y[,i])
  }
  dens = t(dens) # fix dim 
  colnames(dens) = wk_names
  
  
  output = bind_cols(database, dens, avibaseid)
  
  #output fourier smoothed normalized densities
  return(output)
}


#smooth with 8-dim basis
ebird19rel_smoothed = four_smooth(df = ebird19rel, k = 8)
ebird22rel_smoothed = four_smooth(df = ebird22rel, k = 8)
inat19rel_smoothed = four_smooth(df = inat19rel, k = 8)
inat22rel_smoothed = four_smooth(df = inat22rel, k = 8)
```


#save fourier smoothed data
```{r}
# write.csv(ebird19rel_smoothed, "~/Desktop/repos/birds_dash/data/ebird19rel_fsmoothed.csv")
# write.csv(ebird22rel_smoothed, "~/Desktop/repos/birds_dash/data/ebird22rel_fsmoothed.csv")
# write.csv(inat19rel_smoothed, "~/Desktop/repos/birds_dash/data/inat19rel_fsmoothed.csv")
# write.csv(inat19rel_smoothed, "~/Desktop/repos/birds_dash/data/inat22rel_fsmoothed.csv")
```


```{r}
#optional: plot a couple comparisons to ensure fourier step working properly
#yes!

# ebird19rel_smoothed = four_smooth(df = ebird19rel, k = 8)
# 
# for(i in 1:10){
#   plot(1:52, ebird19rel[i,2:53], type = "l"); lines(1:52, ebird19rel_smoothed[i,2:53], col = "red")
# }


```


# create functions for plotting

```{r}

#function for long format
make_long = function(df){
  dflong = df %>%
    pivot_longer(!c(AVIBASEID, yr_db), 
                 names_prefix = "Week-", 
                 names_to = "week",
                 values_to = "count")
  dflong$week = as.numeric(dflong$week)
  #dflong$dataset = deparse(substitute(df))
  return(dflong)
}


ebird19rel_smoothed_long = make_long(ebird19rel_smoothed)
ebird22rel_smoothed_long = make_long(ebird22rel_smoothed)
inat19rel_smoothed_long = make_long(inat19rel_smoothed)
inat22rel_smoothed_long = make_long(inat22rel_smoothed)


allbirdsrel_smoothed_long = rbind(ebird19rel_smoothed_long,
                                  ebird22rel_smoothed_long,
                                  inat19rel_smoothed_long,
                                  inat22rel_smoothed_long)
```


```{r}

#function for plotting relative smoothed curves on the circular/cartesian domains

#right now this function just works on i, the index of the id in the vector which contains all bird ids
#it matches it to the corresponding AVIBASEID and then the common name using bird_dict

ggcirc = function(i, 
                  circular = FALSE, 
                  scalefix = FALSE){
  #base
  gg_base = ggplot(allbirdsrel_smoothed_long %>% 
                     filter(AVIBASEID %in% birdids[i]), 
                   aes(x = week, 
                       y = count, 
                       group = yr_db, 
                       color = yr_db)) + 
    geom_hline(yintercept=0, lwd = 1.25, color = "grey") + 
    geom_hline(yintercept=1/52, color = "grey", linetype="dashed", lwd = 1.25) + 
    ylab("frequency") + 
    annotate("rect", 
             xmin = 1, 
             xmax = 9, 
             ymin = 0, 
             ymax = Inf, 
             fill= "#B983FF", 
             alpha = 0.2) + 
    annotate("rect", 
             xmin = 9, 
             xmax = 9+13, 
             ymin = 0, 
             ymax = Inf, 
             fill= "#00BA38", alpha = 0.2) + 
    annotate("rect", 
             xmin = 22, 
             xmax = 35, 
             ymin = 0, 
             ymax = Inf, 
             fill= "#A3A500", alpha = 0.2) + 
    annotate("rect", 
             xmin = 35, 
             xmax = 48, 
             ymin = 0, 
             ymax = Inf, 
             fill= "#E58700", alpha = 0.2) +
    annotate("rect", 
             xmin = 48, 
             xmax = 52, 
             ymin = 0, 
             ymax = Inf, 
             fill= "#B983FF", alpha = 0.2)
  

  
  
  matchidx = match(birdids[i], bird_dict$AVIBASEID)
  common = bird_dict$CommonName[matchidx]
  
  #add bird curve plots
  if(circular){
    gg_onebird = gg_base +
      geom_polygon(fill = NA, lwd = 1.1) + 
      ggtitle(common) + 
      theme(text=element_text(size=20),
            aspect.ratio = 10/11) + 
      labs(color = "Database/Year") + 
      coord_polar() 
  }else{
    if(scalefix){
      gg_base = gg_base + ylim(c(0,.12))
    }
    
    gg_onebird = gg_base + 
      geom_line(lwd = 1.1) + 
      ggtitle(common) + 
      theme(text=element_text(size=20),
            aspect.ratio = 10/11, 
            legend.position = "none") +
      labs(color = "Database/Year") + 
      guides(group="none", 
             color = "none") + 
      coord_cartesian(expand = FALSE)
  }

  
  gg_onebird
}
```

#function for plotting side by side

```{r}

sidebysideplot = function(i, scalefix = FALSE){
  common = bird_dict$CommonName[match(i, bird_dict$AVIBASEID)]
  
  flat = ggcirc(i, circular = FALSE, scalefix = scalefix)
  circ = ggcirc(i, circular = TRUE, scalefix = scalefix)
  
  gg = flat + circ + theme(plot.margin=grid::unit(c(0,0,0,0), "mm")) + 
    plot_layout(ncol = 2, nrow = 1, guides = "collect")
  
  return(gg)

}

```

```{r, fig.width=14}
sidebysideplot(10, scalefix = TRUE)
```


For Seneth--

What we want next:

Create and host a website/github which allows you to display these side by side plots for a selected bird. Make a dropdown bar of the 260 birds available in this dataset that the user can scroll thru and select. Allow for options to switch scalefix on/off or to compare only specific dbs/years. Allow for settings to plot the unsmoothed curves. 

Potentially allow for the user to select multiple birds at a time to facilitate comparisons. 



